Spatial learning and brain mitochondria are robust to prenatal conditions in the delicate skink

### 2025-03-16 ms.qmd
Updates

---
docx: 
bibliography: "./bib/refs.bib"
csl: "./bib/biology-letters.csl"
reference-doc: "./bib/tmpl.docx"
execute:
  echo: false
  error: false
  cache: false
  warning: false
link-citations: true
crossref:  
  fig-title: Fig    # (default is "Figure")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Table  # (default is "Table")
editor_options: 
  chunk_output_type: console
---

@@

@@

Cognition encompasses processes such as perception, learning, memory, and decision-making, which allow animals to gather, retain, and use environmental information [@shettleworth]. These processes are fundamental for survival and reproduction, influencing behaviours like foraging, mate selection, or antipredatory strategies [@dukas_evolutionary_2004]. Learning - the acquisition and consolidation of new information [@dukas_evolutionary_2004] - enables individuals to create new associations between events, which can be essential for coping with environmental changes [@dukas_evolutionary_2004; @leal_behavioural_2012; @buchanan_condition_2013]. However, the capacity to form these associations varies among individuals, potentially affecting their responses to environmental challenges [@ward2016ecological; @welklin2024spatial]. Animals that learn faster may exploit novel resources better or may start responding to new threats more efficiently than slow learners [@ward2016ecological]. Consequently, individual differences in learning abilities can have profound effects on survival and reproductive output, ultimately influencing population growth rates and stability [@ward2016ecological; @welklin2024spatial].  Understanding the mechanisms that drive these differences is therefore crucial for predicting how populations will respond to novel conditions.

Factors like age, sex, or early- life conditions can have important effects on learning abilities [@szuran_water_1994; @lemaire_prenatal_2000; @zhu_prenatal_2004; @amiel_hotter_2012; @amiel_egg_2014; @carazo_sex_2014; @noble_age-dependent_2014]. Early developmental environment is particularly influential, as the brain is highly sensitive to environmental inputs during this period [@zhu_prenatal_2004]. Early experiences can alter neurotransmitter production [@amani2021perinatal], gene expression [@zhou2020effects], or brain structure [@amiel_effects_2017], with lasting effects on cognition. Among the physiological mechanisms underlying learning, mitochondrial activity has emerged as a key factor influencing brain function [@siegel1994basic; @du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018]. The neural processes involved learning impose substantial energetic demands [@mcnay_decreases_2000; @mann_coupling_2021; @alexandrov_neuronal_2022], making efficient mitochondrial respiration essential [@du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018]. Additionally, learning depends on a dense network of functional neurons [@lefebvre_taxonomic_2011; @amiel_effects_2017], which can be compromised by excessive oxidative stress byproduct of mitochondrial function [@finkel_oxidants_2000; @zhu_prenatal_2004; @du_dynamic_2009; @gong_chronic_2011; @hoffmann_mitochondrion_2018]. Studies on domestic rats (_Rattus norvegicus_) have shown that the effects of early environment on mitochondrial physiology can have profound consequences for cognitive abilities [@zhu_prenatal_2004; @cao2014ampk]. However, the extent to which these effects are generalizable to other taxa, particularly ectotherms, remains largely unexplored.

Since mitochondria are maternally inherited, maternal condition plays a fundamental role in shaping offspring mitochondrial activity [@picard_mitochondria_2014]. In particular, maternal stress can strongly influence how mitochondria operate in the offspring [@zhu_prenatal_2004]. Stressful situations elevate glucocorticoids (GCs) [@sapolsky_how_2000], which can be transmitted to the developing embryo [@uller_sex-specific_2009], altering mitochondrial physiology through transgenerational effects [@picard_mitochondria_2014].For instance, maternal stress has been shown to contribute significantly to oxidative stress in the brain of rats (_Rattus norvegicus_) [@zhu_prenatal_2004; @haussmann_embryonic_2012].

@@

Here, we examined how prenatal temperature and corticosterone (CORT) — the primary GC in reptiles — affect brain mitochondrial physiology and spatial learning in the delicate skink (_Lampropholis delicata_). We hypothesized that prenatal CORT and temperature would influence learning abilities through their effects on mitochondrial activity in the brain. Specifically, we predicted that prenatal CORT would decrease energy production while increasing oxidative damage [@costantini2011meta; @gong_chronic_2011; but see @crino2024eggs]. Similarly, high temperatures would decrease mitochondrial efficienty [@zavorka_climate_2021; @crino2024eggs], but with the potential to reduce oxidative stress [@treidel2016temperature]. We further predicted that the combined effects of CORT and temperature would lead to complex interactions, with both factors negatively impacting mitochondrial efficiency but having opposite effects on oxidative stress. These effects would lead to differences in learning abilities, that could be affected by the balance between energy production and oxidative stress [@du_dynamic_2009; @picard_energetic_2018; @alexandrov_neuronal_2022]. By examining these interactions, we aim to clarify how prenatal environmental factors shape learning abilities through mitochondrial function, shedding light on the role of early-life conditions in cognitive development and its mechanismsm.

@@

#### Animal husbandry  
*Breeding colony* - Juveniles of _L. delicata_ came from a breeding colony established in the laboratory since 2019. This colony consisted of 270 adults housed in plastic containers (41.5 L x 30.5 W x 21 H cm) with six lizards (two males and four females) per enclosure. Enclosures were provided with shelter, nonstick matting, and several small water dishes. The lizards were fed approx. 40 mid-size crickets (_Acheta domestica_) per enclosure three days a week, and water was given daily. The crickets were dusted with calcium weekly and multivitamin and calcium biweekly. Room temperatures were set to 22-24 ºC, but we also provided the enclosures with a heat chord and a heat lamp following a 12 h light:12 h dark cycle keeping warm side of enclosures is usually at 34 ºC.

#### CORT and Temperature manipulation
We tested the combined effects of prenatal CORT and temperature by elevating CORT concentrations in eggs and then incubating them under one of two temperature regimes (Cold - 23 ± 3 ºC or Hot - 28 ± 3 ºC) in a 2x2 factorial design (@fig-Methods **LTTER**). CORT-treated eggs were topically supplied with 5 µL of crystalline corticosterone (Sigma, Cat. No. C2505) dissolved in 100% ethanol at a final 10 pg CORT/mL concentration (CORT treatment), while Control eggs received an equal volume of 100% Ethanol. CORT dose increased mean yolk CORT levels ~3.7x higher than control eggs in previous studies [@crino2024eggs]. Eggs were then incubated in one of the two previously mentioned temperature regimes that are within the natural limits in _L. delicata_ [@cheetham2011embryonic].

We counterbalanced the number of eggs per clutch assigned to each hormone and temperature using a partial split-clutch design where at least one egg per clutch was assigned to each treatment. For clutches bigger than four, we assigned the remaining eggs randomly to one of the treatments. When we found less than four eggs in a clutch, we assigned each egg randomly to one of the treatments.

#### Learning  
The spatial learning task involved training lizards to navigate a 6-arm maze to reach an exit connected to a transport box that allowed us to return the lizards to their enclosure without further contact (see @fig-Methods **LTTER**). In each trial, the lizards were placed by hand in the center of maze and left to acclimate for two minutes (see @fig-Methods **LTTER**). During acclimatation, the central are was surrounded by a yellow device mounted on a pulley system. At the start of each trial, this device was lifted to startle the lizard. If the lizard did not immediately choose an arm, it was gently prodded with a brush at the end of the tail. Once the lizard made a choice, the brush was used to encourage movement but without guiding it towards any specific arm. If the lizard did not choose correctly after 20 errors — at which point it typically ceased responding — the lizard was gently guided to the correct arm.

Extra-maze cues (but not intra-maze cues) were available to assist with navigation (see @fig-Methods **LTTER**). This design aimed to assess the lizards' ability to learn spatial relationships based on external cues rather than relying on familiar objects within the maze. Additionally, to prevent lizards from relying on subtle intra-maze cues, the maze was replaced every three trials with one of four identical mazes. Each replacement preserved the correct arm’s orientation and the maze’s position within the room, ensuring consistency for each individual. The maze was also cleaned with 70% ethanol between trials to avoid the influence of chemical cues. To control for potential side biases, the correct arm was randomly assigned to one of the six arms for each trial. We employed four different maze orientations (North-West, South-West, South, and South-East), and the number of lizards assigned to each orientation was counterbalanced across treatments.

The task was repeated once daily for 40 consecutive days, and the number of mistakes made by each lizard was recorded. A mistake was defined as the lizard inserting its head into one of the incorrect arms. The primary measure of spatial learning was the total number of errors made before the lizard chose the correct arm.

```{r, fig-Methods}
#| label: fig-Methods
#| fig.cap: "Panel **A** shows the experimental design of the study. On top, the treatments applied to the eggs. On the bottom the learning task and the brain region extracted together with the physiological analyses performed. In panel **B** the details and measurements of the spatial learning maze."

knitr::include_graphics("./Others/SPAL_METH.svg")

#### Brain mitochondrial activity
Immediately following the completion of the tests, we asessed mitochondrial physiology in the brain of all remaining lizards (see below). We euthanized lizards via peritoneal injection 10 mg/kg of a 10 mg/mL alfaxan solution (a potent anesthetic), followed by decapitation. Before decapitation we evaluated the lizard’s response by turning it upside down and by testing the pinching relfex in one of the front limbs. This protocol was approved by the Animal Ethics Committee of the Australian National University (Protocol number: A2022_33).

After decapitation, we extracted manually the medial cortext in the telencephalon (@fig-Methods **LETTER**). We selected the medial cortex because is considered homologous to the mammalian hippocampus, where spatial cognition is encoded [@rodríguez2002conservation; @naumann2015reptilian]. The tissue was transferred to 1xPBS solutioin and then homogenized mechanically using a 100 µm mesh filter (pluriStrainer). The resultant homogenate was divided into two aliquotes: one was used fresh for mitochondrial density, membrane potential, ROS assays; and the other was cryopreserved for later measurements of oxidative damage (DNA damage, lipid peroxidation).

Fresh homogenate suspensions were stained with 5 µM MitoTracker Deep Red FM, 2.5 µM MitoTracker Orange CMTMRos, and 50 µM MitoSOX Red. We used these fluorescent probes to assess mitochondrial density, mitochondrial membrane potential, and superoxide (ROS) production, respectively. We also added 5 µL of #µM Hoechst 33342 Nuclear Viability Dye to each sample to differentiate viable cells from debris. After adding all the dyes, the samples were incubated at 32 °C for 30 minutes and then immediately transferred to flow cytometry facilities for data collection, within 2 hours.

The aliquotes reserved for oxidative damage assays, were stained with 20 µL of #µM Hoechst 33342 Nuclear Viability Dye and 20 µL of 100 µM BODIPY 665/676 Lipid Peroxidation Sensor before cryopreservation. These dyes were used to measure cell viability and lipid peroxidation, respectively. The sampled were then fixed in 1% Neutral-Buffered Formalin, washed, and then preserved in a 1X Tris-EDTA solution with 10% DMSO at -20 °C. On the day of the oxidative damage assays, the samples were thawed, and DMSO was removed. Then, cell membranes were permeabilized by incubating the samples in 200 µL of 1X PBS containing 5mM (6.14 mg/mL) digitonin for 20 minutes at 32°C. Following membrane permeabilization we stained the samples with 20µL of 70µM 8-OHdG Polyclonal Antibody, and allowed them to incubate overnight (~12 hours). 8-OHdG Polyclonal Antibody was used to assessed 8-OHdG, a marker of oxidative damage on DNA. The following day we counterstained the cells with 20µL of #µM (100µg/mL) H+G Goat Anti-Rabbit Conjugate Antibody with Alexa-Fluor 488 at 32°C for 20 minutes and performed the flow cytometry. Oxidative damage assays were performed within 6 months of the initial analysis of fresh samples.

Flow cytometry assays were performed on a Becton Dickson LSRFortessa X-20 flow cytometer using the default wavelength filters. The detectors and voltage settings for each assay were determined during pilot trials and and remained consistent throughout the experiment to ensure comparability across different plates and samples. Data was processed using FlowJo (v. 10.1) software. We used the geometric mean as our main response variables in analyses. For further details on the homogenization, staining, or flow cytometry assays, see the Supplementary Material.

#### Statistical analyses

We performed all the analyses using brm models from stan [@stan] on R version 4.4.0 [@R]. We run a series of univariate and miltivariate models to test the effects of early envirnoment on each variable separately plus the relationships between physiology and learning. All models consisted of four MCMC chains of 8000 iterations, with a warm up interval of 2000 iterations.

Univariate models were used to test the effects of early environment on each of the variables recorded: mitochondrial density, mitochondrial potential, ROS, DNA damage, lipid peroxidation, and the number of errors as a measure of learning. We included in all the models the hormone treatment (CORT vs Control), incubation temperature (Cold vs Hot) and their interaction. Sex and age were included in preliminary models, and excluded from the final models when they were not significant (see Supplementary Material). For all univariate models, clutch identity was included as a random factor. Mitochondrial related variables (density, potential, ROS, DNA damage, and lipid peroxidation) where log-transformed when necessary and all were mean-centered and standardized by dividing them by two times the standard deviation [**REF DAN**]. These variables were considered to follow a normal distribution. The number of errors was modeled as a function of trial, cort, temperature, and the triple interaction. We included in this model the trial within each level of lizard identity as a random slope. The error structure was modelled using a negative binomial distribution with a logit link function [negbinomial(link = "log")]. Otherwise, the procedure was as in the other models.

We used the posterior distributions of parameters from these models to test for differences between treatments. Learning slopes were obtained using the ‘trial’ estimates and its interaction with hormone and temperature treatments. Slope estimates >0 were considered as evidence of learning. pMCMC were used to test the hypothesis that slopes and slope contrasts are different from zero. We considered an effect statistically significant when pMCMC <0.05.

To explore links between mitochondrial function and detection latency, we fitted a multivariate model on brm. We included in the model all the variables of interest and their interactions structured following a specific hypothesis (see @fig-SEM). Individual measurements for learning performance were obtained by modelling errors as a function of trial. We included the trial within each level of lizard identity as a random slope, which was used to obtain the learning slopes for each individual. These learning slopes, as well as the rest of variables, were standardized as before (see above). We assumed a gaussian family for all the variables. Factors found as non-significant in the univariate models were excluded. Clutch identity remained a random factor. Direct, indirect, and total effects were derived from posterior estimates following a Structural Equations Modelling approach [@kline2005principles] (see **TABLE** in Supplementary Material).

@@

**Originally, we started with 96 lizards, 48 per species and 12 per treatment per species. However, due to natural mortality (n = 11), no completion of the training stage (n = 1), or no motivation during the learning tasks (n = 3; see above), we ended up with a total of 81 lizards. Final sample sizes per treatment and species are disclosed on @tbl-data.**

@@

@@

>>>>

---
title: "Spatial learning and brain mitochondria are robust to prenatal conditions in the delicate skink"
docx: 
bibliography: "./bib/refs.bib"
csl: "./bib/biology-letters.csl"
reference-doc: "./bib/tmpl.docx"
execute:
  echo: false
  error: false
  cache: false
  warning: false
link-citations: true
crossref:  
  fig-title: Fig    # (default is "Figure")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Table  # (default is "Table")
editor_options: 
  chunk_output_type: console
---

@@

Pablo Recio^1^,$\ddagger$ , Dalton C. Leibold^1^, Ondi L. Crino ^1,2^, Christopher R. Friesen^3^, Daniel W.A. Noble^1^

^1^ Division of Ecology and Evolution, Research School of Biology, The Australian National University, Canberra, ACT 2601, Australia  
^2^ Flinder's University, College of Science and Engineering, Bedford Park, SA 5042, Australia  
^3^ University of Wollongong, Wollongong, NSW 2500, Australia  
$\ddagger$ Corresponding author: pablo.reciosantiago@anu.edu.au

ORCID:

Pablo Recio ORCID: 0000-0002-5890-0218  
Dalton C. Leibold ORCID: 0000-0001-9645-2033  
Ondi L. Crino ORCID: 0000-0001-5700-1387  
Christopher R. Friesen ORCID: 0000-0001-5338-7454  
Daniel W.A. Noble ORCID: 0000-0001-9460-8743

## Abstract
Learning is highly sensitive to environmental conditions, including early thermal environments and prenatal exposure to glucocorticoids. These factors influence mitochondrial activity, which plays a crucial role in neural development and synaptic plasticity, ultimately shaping cognitive abilities. Here, we investigated the interactive effects of prenatal temperature and corticosterone (CORT) — the primary glucocorticoid in reptiles — on spatial learning and brain mitochondrial activity in the delicate skink (_Lampropholis delicata_). We supplemented eggs with either CORT or a control vehicle and incubating them under two thermal regimes: cold (23 ± 3°C) or hot (28 ± 3°C). Juveniles were then tested in a spatial learning task over 40 days, after which mitochondrial activity in the medial cortex was assessed. Our results revealed that both mitochondrial physiology and spatial learning in _L. delicata_ are robust to prenatal temperature and CORT exposure. In addition, we found no significant relationship between mitochondrial function and cognitive abilities. However, mitochondrial potential increased ROS levels, which could impact oxidative stress under high metabolic demands or low antioxidant capacity. These findings highlight the cognitive and physiological resilience of _L. delicata_ to early-life environmental challenges. Future research should explore whether this robustness extends to other brain regions and life stages.

@@

Learning - the acquisition and consolidation of new information [@dukas_evolutionary_2004] - enables individuals to create new associations between events, which can be essential for coping with environmental changes [@dukas_evolutionary_2004; @leal_behavioural_2012; @buchanan_condition_2013]. However, the capacity to form these associations varies among individuals, potentially affecting their responses to environmental challenges [@ward2016ecological; @welklin2024spatial]. Animals that learn faster may exploit novel resources better or may start responding to new threats more efficiently than slow learners [@ward2016ecological]. Consequently, individual differences in learning abilities can have profound effects on survival and reproductive output, ultimately influencing population growth rates and stability [@ward2016ecological; @welklin2024spatial]. Understanding the mechanisms that drive these differences is therefore crucial for predicting how populations will respond to novel conditions.

Factors like age, sex, or early- life conditions can have important effects on learning abilities [@szuran_water_1994; @lemaire_prenatal_2000; @zhu_prenatal_2004; @amiel_hotter_2012; @amiel_egg_2014; @carazo_sex_2014; @noble_age-dependent_2014]. Early developmental environment is particularly influential, as the brain is highly sensitive to environmental inputs during this period [@zhu_prenatal_2004]. In fact, early experiences can alter neurotransmitter production [@amani2021perinatal], gene expression [@zhou2020effects], or brain structure [@amiel_effects_2017], with lasting effects on cognition. Among the physiological mechanisms underlying learning, mitochondrial activity is considered a key factor [@siegel1994basic; @du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018]. The neural processes involved learning impose substantial energetic demands [@mcnay_decreases_2000; @mann_coupling_2021; @alexandrov_neuronal_2022], making efficient mitochondrial respiration essential [@du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018]. Additionally, learning depends on a dense network of functional neurons [@lefebvre_taxonomic_2011; @amiel_effects_2017], which can be compromised by excessive oxidative stress byproduct of mitochondrial function [@finkel_oxidants_2000; @zhu_prenatal_2004; @du_dynamic_2009; @gong_chronic_2011; @hoffmann_mitochondrion_2018]. Studies on mammals have shown the pervassive effects of mitochondrial physiology on cognitive abilities [@zhu_prenatal_2004; @hara_presynaptic_2014; @cao2014ampk]. However, the extent to which these effects are generalizable to other taxa remains largely unexplored.

Since mitochondria are maternally inherited, maternal condition plays a fundamental role in shaping offspring mitochondrial activity [@picard_mitochondria_2014]. In particular, maternal stress can strongly influence how mitochondria operate in the offspring [@zhu_prenatal_2004]. Stressful situations elevate glucocorticoids (GCs) [@sapolsky_how_2000], which can be transmitted to the developing embryo [@uller_sex-specific_2009] altering mitochondrial physiology through transgenerational effects [@picard_mitochondria_2014]. For instance, maternal stress has been shown to contribute significantly to oxidative stress in the brain of rats (_Rattus norvegicus_) [@zhu_prenatal_2004; @haussmann_embryonic_2012; @cao2014ampk].

@@

Here, we examined how prenatal temperature and corticosterone (CORT) — the primary GC in reptiles — affect brain mitochondrial physiology and spatial learning in the delicate skink (_Lampropholis delicata_). We hypothesized that prenatal CORT and temperature would influence learning abilities through impacts on brain mitochondrial activity. Specifically, we predicted that prenatal CORT would decrease energy production while increasing oxidative damage [@costantini2011meta; @gong_chronic_2011; but see @crino2024eggs]. Similarly, high temperatures would decrease mitochondrial efficienty [@zavorka_climate_2021; @crino2024eggs], but with the potential to reduce oxidative stress [@treidel2016temperature]. We further predicted that the combined effects of CORT and temperature would lead to complex interactions, with both factors negatively impacting mitochondrial efficiency but having opposite effects on oxidative stress. These effects would lead to differences in learning abilities, that could be affected by the balance between energy production and oxidative stress [@du_dynamic_2009; @picard_energetic_2018; @alexandrov_neuronal_2022]. By examining these interactions, we aim to clarify how prenatal environmental factors shape learning abilities through mitochondrial function, shedding light on the role of early-life conditions in cognitive development and its mechanismsm.

@@

#### Animals' origin and collection
Subjects came from a breeding colony established in the laboratory since 2019. It consisted in 270 adults living in groups of two males and four females. Eggs were collected from these groups between mid-Novemnber 2023 to mid-January 2024. After collecting the eggs, we treated them with CORT or vehicle and incubated them under two different temperature regimes (see below). Clutch and egg identity were asigned immediately after collection, and the eggs were incubated in individual cups until hatching. Hatchlings were also kept in individual enclosures until the end of the experiment. For details on husbandry and breeding conditions, see Methods 1: Animal husbandry in Supplementary Material.

#### CORT and Temperature manipulation
We tested the combined effects of prenatal CORT and temperature by elevating CORT concentrations in eggs and then incubating them under one of two temperature regimes (Cold - 23 ± 3 ºC or Hot - 28 ± 3 ºC) in a 2x2 factorial design. We used a partial split clutch design where eggs from a given clutch were distributed equally across the four treatments when clutch sizes were larger than four and randomly across treatments when less than four. CORT-treated eggs were topically supplied with 5 µL of crystalline corticosterone (Sigma, Cat. No. C2505) dissolved in 100% ethanol at a final 10 pg CORT/mL concentration (CORT treatment), while Control eggs received an equal volume of 100% Ethanol. CORT dose increased mean yolk CORT levels ~3.7x higher than control eggs in previous studies [@crino2024eggs]. Eggs were then incubated in one of the two previously mentioned temperature regimes that are within the natural limits in _L. delicata_ [@cheetham2011embryonic].

#### Learning  
The spatial learning task involved training lizards to navigate a 6-arm maze to reach an exit connected to a transport box that allowed us to return the lizards to their enclosure without further contact. In each trial, the lizards were placed by hand in the center of maze and left to acclimate for two minutes. During acclimatation, the central area was surrounded by a yellow device mounted on a pulley system. At the start of each trial, this device was lifted to startle the lizard. If the lizard did not immediately choose an arm, it was gently prodded with a brush at the end of the tail. Once the lizard made a choice, the brush was used to encourage movement but without guiding it towards any specific arm. If the lizard did not choose correctly after 20 errors — at which point it typically ceased responding — the lizard was gently guided to the correct arm.

We assessed lizards' spatial learning using external cues for navigation while avoiding intra-maze cues. To avoid subtle intra-maze cues, the maze was replaced every three trials with one of four identical mazes. Each replacement preserved the correct arm’s orientation and the maze’s position within the room, ensuring consistency for each individual. Additionally, the maze was cleaned with 70% ethanol between trials to avoid the influence of chemical cues. To control for potential side biases, the correct arm was randomly assigned to one of the six arms for each trial. We employed four different maze orientations (North-West, South-West, South, and South-East), and the number of lizards assigned to each orientation was counterbalanced across treatments.

The task was repeated once daily for 40 consecutive days, and the number of errors made by each lizard was recorded. An error was defined as the lizard inserting its head into one of the incorrect arms, choosing the same arm two or more times consecutively was counted as independent errors.

#### Brain mitochondrial activity
Immediately following the completion of the tests, we asessed mitochondrial physiology in the brain of all remaining lizards (see below). We euthanized lizards via peritoneal injection 10 mg/kg of a 10 mg/mL alfaxan solution (a potent anesthetic), followed by decapitation. Before decapitation we evaluated lizard's righting response and the pinching reflex in one of the front limbs.

After decapitation, we extracted the medial cortext in the telencephalon. We selected the medial cortex because is considered homologous to the mammalian hippocampus, where spatial cognition is encoded [@rodríguez2002conservation; @naumann2015reptilian]. The tissue was transferred to 1xPBS solution and then homogenized mechanically using a 100 µm mesh filter (pluriStrainer). The resultant homogenate was divided into two aliquotes: one was used fresh for mitochondrial density, membrane potential, and ROS assays; and the other was cryopreserved for later measurements of DNA damage and lipid peroxidation.

Fresh homogenate suspensions were stained with 5 µM MitoTracker Deep Red FM, 2.5 µM MitoTracker Orange CMTMRos, and 50 µM MitoSOX Red. We used these fluorescent probes to assess mitochondrial density, mitochondrial membrane potential, and superoxide (ROS) production, respectively. We also added 5 µL of #µM Hoechst 33342 Nuclear Viability Dye to each sample to differentiate viable cells from debris. These samples were analyzed in a flow cytometer within 2 hours of brain extraction.

The aliquotes reserved for oxidative damage assays were stained with 20 µL of #µM Hoechst 33342 Nuclear Viability Dye and 20 µL of 100 µM BODIPY 665/676 Lipid Peroxidation Sensor before cryopreservation. These dyes were used to measure cell viability and lipid peroxidation, respectively. The samples were then fixed in 1% Neutral-Buffered Formalin, washed, and preserved in a 1X Tris-EDTA solution with 10% DMSO at -20 °C. On the day of the oxidative damage assays, the samples were thawed, and DMSO was removed. Then, cell membranes were permeabilized in 200 µL of 1X PBS containing 5mM (6.14 mg/mL) digitonin. Following membrane permeabilization we stained the samples with 20µL of 70µM 8-OHdG Polyclonal Antibody - a marker of DNA damage - and allowed them to incubate overnight (~12 hours). The following day we counterstained the cells with 20µL of #µM (100µg/mL) H+G Goat Anti-Rabbit Conjugate Antibody with Alexa-Fluor 488 and analyzed the samples in the the flow cytometer. Oxidative damage assays were performed within 6 months of the initial analysis of fresh samples.

Flow cytometry was performed on a Becton Dickson LSRFortessa X-20 flow cytometer using the default wavelength filters. The detectors and voltage settings for each assay were determined during pilot trials and and remained consistent throughout the experiment. Data was processed using FlowJo (v. 10.1) software. We obtained the mean fluorescent intensity for mitochondrial density, mitochondrial potential, ROS production, DNA damage, and lipid peroxidation. For further details on the homogenization, staining, or flow cytometry assays, see Methods 2: flow cytometry in Supplementary Material.

#### Statistical analyses
We performed all the analyses using brm models from stan [@stan] on R version 4.4.0 [@R]. We run a series of univariate and multivariate models to test the effects of early envirnoment on each variable separately plus the relationships between physiology and learning. All models consisted of four MCMC chains of 8000 iterations, with a warm up interval of 2000 iterations.

Univariate models were used to test the effects of early environment on each of the variables recorded: mitochondrial density, mitochondrial potential, ROS, DNA damage, lipid peroxidation, and the number of errors as a measure of learning. We included in all the models the hormone treatment (CORT vs Control), incubation temperature (Cold vs Hot) and their interaction. Sex and age were included in preliminary models and excluded from the final models when they were not significant. For all univariate models, clutch identity was included as a random factor. Mitochondrial related variables (density, potential, ROS, DNA damage, and lipid peroxidation) where log-transformed when necessary and all were mean-centered and standardized by dividing them by two times the standard deviation [@gelman2008scaling]. These variables were considered to follow a normal distribution.

The number of errors was modeled as a function of trial, cort, temperature, and the triple interaction. We included in this model the trial within each level of lizard identity as a random slope. The error structure was modelled using a negative binomial distribution with a logit link function [negbinomial(link = "log")]. Otherwise, the procedure was as in the other models.

We used the posterior distributions of parameters from these models to test for differences between treatments. Learning slopes were obtained using the ‘trial’ estimates and its interaction with hormone and temperature treatments. Slope estimates <0 were considered as evidence of learning. pMCMC were used to test the hypothesis that slopes and slope contrasts are different from zero. We considered an effect statistically significant when pMCMC <0.05.

To explore links between mitochondrial function and detection latency, we fitted a multivariate model on brm. We included in the model all the variables of interest and their interactions structured following a specific hypothesis (see @fig-sem). Individual measurements for learning performance were obtained by modelling errors as a function of trial. We included the trial within each level of lizard identity as a random slope, which was used to obtain the learning slopes for each individual. These learning slopes, as well as the rest of variables, were standardized as before (see above). We assumed a gaussian family for all the variables. Factors found as non-significant in the univariate models were excluded. Clutch identity remained a random factor. Direct, indirect, and total effects were derived from posterior estimates following a Structural Equations Modelling approach [@kline2005principles] (see Table S4).

One of the plates used to analyze oxidative damage malfunctioned, compromising data quality. This data was excluded from the final analysis and, as a result, the sample size for DNA damage and lipid peroxidation is smaller than for the other variables. For the multivariate model, we imputed the missing values. The final sample sizes are presented below.

@@

We started with 80 lizards - 20 per treatment - coming from `r total_clutches` clutches. However, due to natural mortality (n = 2), only 79 lizards were included in the learning analyses, and 78 in the mitochondrial analyses. Besides, due to malfunctioning of the flow cytometer, we ended up with 40 inividuals (10 per treatment) for DNA damage and lipid peroxidation. Final sample sizes per treatment and species are disclosed on @fig-learning and @fig-results_mit.

@@

```{r, fig-learning}
#| label: fig-learning
#| fig.cap: "Results for learning analyses. Panel A shows the predicted number of errors over trials. The lines represent the mean predicted number of errors for each trial, and the shaded areas indicate the standard deviation of the mean; both were obtained using the slope and intercept estimates from the posterior distributions. Panel B shows the distribution of the estimates of slopes per each treatment. The x-axis represents the slope estimate, and in the y-axis are the density of the estimates. Points and bars represent the mean and standard deviation of the estimated slopes, respectively. Dashed lines indicate value 0. The different colors in both panels indicate the different treatments."
source(here("R", "func.R"))
#
#
####### A) Create df
# Slopes df (employed in fig-learning B)
slope_list <- list(`Control-Cold` = slope_ControlCold,
                 `CORT-Cold` = slope_CORTCold,
                 `Control-Hot` = slope_ControlHot,
                 `CORT-Hot` = slope_CORTHot)
data_fig_learning_slopes <- do.call(rbind, lapply(names(slope_list), function(x) {
  data.frame(treatment = x, slopes = slope_list[[x]])
}))
# Intercepst df (only for fig-learning A)
intercept_list <- list(`Control-Cold` = int_ControlCold,
                 `CORT-Cold` = int_CORTCold,
                 `Control-Hot` = int_ControlHot,
                 `CORT-Hot` = int_CORTHot)
data_fig_learning_intercepts <- do.call(rbind, lapply(names(intercept_list), function(x) {
  data.frame(treatment = x, intercepts = intercept_list[[x]])
})) %>% 
  group_by(treatment) %>%
  summarize(int_sd = sd(intercepts),
            intercepts = mean(intercepts))
# df fig-learning A
data_fig_learning_slopesA <- data_fig_learning_slopes %>%
  group_by(treatment) %>%
  summarize(slope_sd = sd(slopes),
            slopes = mean(slopes),)
data_fig_learningA <- merge(data_fig_learning_slopesA, data_fig_learning_intercepts, by = "treatment")
treatment <- unique(data_fig_learningA$treatment)
fig_A_df <- data.frame()
for(t in treatment){ # Loop per treatment
  df <- data_fig_learningA %>%
    filter(treatment == t) %>%
    data.frame()
  # Variables selected
  m <- df$slopes
  u <- df$intercepts
  # Loop per treatment
  num_individuals <- length(u)
  for(x in 0:40){
    if (x == 0){
      sd <- df$int_sd
    } else {
      q25 <- df$slope_sd
    }
    value <- exp(u + m * x)
    temp_df <- data.frame(trial = rep(x, length(value)),
                          treatment = rep(t, length(value)),
                          errors = value,
                          sd = sd * value)
    fig_A_df <- rbind(fig_A_df, temp_df)
  }
}
# 
#### C) Plot the fig-learning A and fig-learning B separately 
fig_learningA <- plot_errorsday(fig_A_df)
fig_learningB <- plot_slopes(data_fig_learning_slopes)
#
#### D) Combine plots A and B to have fig-learning
fig_results_learning <- plot_grid(fig_learningA, fig_learningB, nrow = 1) +
  # Insert title for each plot
  annotate("text", x = 0.045, y = 0.935, label = "A", hjust = 1, vjust = 1, size = 7, fontface = "bold") +
  annotate("text", x = 0.545, y = 0.935, label = "B", hjust = 1, vjust = 1, size = 7, fontface = "bold") 
#
ggsave("./output/figures/text/fig_results_errors.png", plot=fig_results_learning , width = 21, height = 7, units = "cm", dpi = 600)
knitr::include_graphics("./output/figures/text/fig_results_errors.png")
```

We found that the lizards improved their choices over time, making less mistakes as the trials progressed (see @fig-learning and Table S1). However, the learning rate was not influenced by CORT, temperature, or their interaction (see contrasts in Table S2).

Mitochondrial density, mitochondrial potential, ROS, DNA damage, and lipid peroxidation was not affected by prenatal conditions (see @fig-results_mit and Table S2). We did not see any significant effects of sex and age on mitochondrial physiology either (see Table S3 and Tables S5-S9).

@@

We found that lizards improved their performance over time, indicating learning. However, this improvement was not influenced by prenatal conditions. Similarly, mitochondrial physiology was not affected by prenatal conditions, and we found no significant relationships between mitochondrial physiology and learning abilities. However, ROS production increased with mitochondrial potential, suggesting that mitochondrial potential may be a good predictor of ROS production in medial cortex of _L. delicata_ with potential implications for the balance between energy production and oxidative damage.

#### Spatial learning in _Lampropholis delicata_ is robust to prenatal conditions  
Lizards exhibited learning, but learning rates were not influenced by prenatal CORT or temperature. This contrasts with findings in other taxa, where incubation temperature and prenatal GC elevations affected learning and brain development [@lemaire_prenatal_2000; @zhu_prenatal_2004; @amiel_hotter_2012; @amiel_effects_2017; @dayananda_incubation_2017]. Our results suggest that _L. delicata_ buffers cognitive functions against early environmental inputs, consistent with previous findings showing that behavioural flexibility in this species is also unaffected by prenatal CORT or temperature [@recio2025early]. However, since prenatal CORT and temperature can have region-specific effects on brain development [@coomber_independent_1997; @sakata_neural_2000], other cognitive domains might be more sensitive to prenatal conditions. Future research should explore whether different brain regions in _L. delicata_ vary in their resilience to prenatal inputs to gain a more comprehensive understanding of cognitive development in this species.

#### Mitochondrial physiology is not affected by prenatal conditions  
Contrary to our predictions, prenatal temperature and CORT did not significantly influence mitochondrial physiology in the medial cortex of _L. delicata_. Studies in other taxa have been shown that incubation at high temperatures could decrease energy production and oxidative stress [@treidel2016temperature; @zavorka_climate_2021] while, GCs elevation or maternal stress is related to lower mitochondrial efficiency and higher oxidative stress [@costantini2011meta; @gong_chronic_2011; but see @crino2024eggs].
